---
title: "day_2"
author: "Tim"
date: "2025-05-07"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Day 2 Script
```{r}
a <- c(1:15)
b <- matrix(data = 1:12, ncol =2)
c <- array(data = 1:24, dim = c(2,3,2))
```
## Exercise 1
Save all objects in your workspace to an .RData file.
```{r}
save(a,b,c, file = "r_day2_objects.RData") 
#or
save.image(file = "r_day2_objects.RData")
```
Write one object in your workspace to a file using saveRDS().
```{r}
saveRDS(object = a, file = "r_day2_object_a.RData")
```
Remove one object from your workspace.
Prove that the object was removed.
```{r}
rm(b)
ls()
```
Remove all objects from your workspace.
```{r}
rm(list = ls())
```
Display your working directory.
```{r}
getwd()
```
Create a new directory and set the working directory in that new directory.
```{r}
setwd('/project/clme2295/Week 1/')
```
Restore objects saved in the .RData file to your workspace.
```{r}
load('/project/clme2295/r-week-1/r_day_2/r_day2_objects.RData')
```
Restore the object saved in the RDS file to your workspace under a diﬀerent name
```{r}
new_a <- readRDS("r_day2_object_a.RData")
#reset the working directory
setwd('/project/clme2295/r-week-1/r_day_2/')
```


## Exercise 2
Use readRDS() to load the file my_day2matrix.rds and assign the object
to the name m.

```{r}
m <- readRDS("/project/shared/r/1_r_data_science/2-base/my_day2matrix.rds")

```

Compute the sum of values in each row and add those values as a new
column in the matrix.

```{r}
m1 <- cbind(m, rowSums(m))
m1
```

Run the command data("ToothGrowth") to load the builtin data set
ToothGrowth. Open the help page for the ToothGrowth data set, to learn
more about it.

```{r}
data("ToothGrowth")
help("ToothGrowth")
```

What is the class of the ToothGrowth object? **Answer - A dataframe**

What type of data is stored in each column of the ToothGrowth data set?
**Answer - Numeric and factor data**

What is the mean tooth length across all observations in the data set?

```{r}
colnames(ToothGrowth)
Tooth_len <- ToothGrowth$len
mean(Tooth_len)
#or
mean(ToothGrowth$len)
```

What is maximum value of tooth length?

```{r}
max(Tooth_len)
```

What is minimum value of tooth length?

```{r}
min(Tooth_len)
```

Can you use the functions rowSums() and colSums() on the ToothGrowth
object? no as it is is not all numeric data

##Exercise 3 - Sorting Data Frames Load the airquality data set.

```{r}
data("airquality")
```

Open the help page for this data set.

```{r}
help("airquality")
```

Examine the data set.

```{r, results='hide'}
airquality
```

Display the column names of the airquality data frame.

```{r}
colnames(airquality)
```

Sort the data frame by increasing value in the Ozone column.

```{r}
airquality[order(airquality$Ozone), ]
```

Sort the data frame by Month in increasing order and Temp in decreasing
order.

```{r}
airquality[
  order(airquality$Month, airquality$Temp, decreasing = TRUE), 
  ]
```

Write the latest sorted data frame to a text file format of your choice.

```{r}
sorted_airquality <- airquality[order(airquality$Month, airquality$Temp, decreasing = c(TRUE, FALSE)), ]
write.csv(sorted_airquality, file = "sorted_airquality.csv")
```

## Merging Exercise

```{r}
buildings <- data.frame(
site = c(1, 2, 3),
name = c("b1", "b2", "b3")) 
survey_data <- data.frame(
survey = c("A", "A", "A", "B", "B", "B"),
location = c(1, 2, 3, 2, 3, 1),
efficiency = c(51, 64, 70, 71, 80, 58))
```

What is the shared information in these two data frames? **Answer - site
and location**

Use the merge() function to combine the two data frames by the shared
information into a new data frame called buildings_survey.

```{r}
buildings_survey <- merge(x= buildings, y= survey_data, by.x = "site", by.y = "location")
buildings_survey
```

##Summarising groups of data Compute the mean of each numeric column
each month in the airquality data frame using aggregate().

```{r}
aggregate(x=airquality, by = list(month = airquality$Month), FUN = mean)
```

Make sure NA values are removed.

```{r}
aggregate(x=airquality, by = list(month = airquality$Month), FUN = mean, na.rm = TRUE)
```

Compute the mean of the Solar.R column each month.

```{r}
aggregate(x=airquality[, c("Solar.R","Month")], by = list(month = airquality$Month), FUN = mean, na.rm=TRUE)
```

##Exercise - Writing a function Write a function to calculate the
hypotenuse of a triangle given the length of the other two sides.

```{r}
hypotenuse <- function(a,b){
  h <- sqrt((a^2)+(b^2))
  return(h)
}
```

*note that the h in this line of code is only for the purpose of the
function it does not generate a data point. It is ess relvant in a
cingle line of code*

Run the function you have created with different values.

```{r}
hypotenuse(2,3)
```

##Exercise - Apply

Create a vector of integers from 1 to 10.

```{r}
v <- 1:10
```

Compute the log2 of each value in the vector using either lapply() or
sapply().

```{r}
lapply(v, FUN = log2)
```

```{r}
sapply(v, FUN = log2)
```

Create a list of four elements, each element being a vector of type
either numeric or logical.

```{r}
l <- list(1:12, c(TRUE, FALSE, TRUE, TRUE), c(1.23455,3.5633223,56.4564,0.323425), c(TRUE,FALSE))
```

Compute the sum of each vector in the list using either lapply() or
sapply().

```{r}
sapply(l[c(1, 3)], FUN = sum)
```

Part 3 Use sapply() on the list that you created in part 2, to repeat
each element of each vector three times. i.e., 1, 2, 3 should become 1,
1, 1, 2, 2, 2, 3, 3, 3

```{r}
help (rep)
sapply(X=l, FUN = rep, each = 3)
```

##Exercise - Loops and conditions Write a for loop that iterates over
the integers 1 to 7 and prints the number raised to the power of three.

```{r}
for (i in 1:7) {
print(i^3)
}
```

Write a for loop that iterates over the names of the columns in the
builtin data set iris and prints each column name together with the
number of characters in that column name. Example output: Sepal.Length:
12 Hint: use the functions print(), paste0(), and nchar(). Remember to
read the help page of each function to learn more about them.

```{r}
iris_cols <- colnames(iris)
for (column_name in colnames(iris)) { 
  print(paste0(column_name, " : ", nchar(column_name)))
  }
```

Use the ifelse() function to print the name of colours that are made up
of four characters in the vector my_colours below.

```{r}
my_colours <- c("red", "orange", "purple", "yellow", "pink", "blue")
ifelse(test = nchar(my_colours)==4, yes = my_colours, no = ".")

```
